---
title: "MCP Test: Power and Appropriateness"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ECxHelper)
library(PMCMRplus)
library(multcomp)
library(tidyverse)
```

## Power Simulation

```{r}
cv <- 0.1
N <- 10
sigma <- 0.1
mu <- 1*c(1,0.9,0.8,0.7)
dat <- as.data.frame( t(plyr::ldply(mu,function(x){rnorm(N,x,sigma)})))
names(dat) <- c("C","T1","T2","T3")

dat <- pivot_longer(dat,cols=everything()) %>% mutate(name=factor(name))
plot(dat$value~dat$name)
res1 <- williamsTest(dat$value~dat$name,alternative = "less")
getwilliamRes(res1)
jonckheereTest(dat$value~dat$name)
res2 <- stepDownTrendTest(value~name,data=dat,test="jonckheereTest",alternative = "less")
res2 <- stepDownTrendTest(value~name,data=dat,test="jonckheereTest")
summary(res2)
```


```{r}
sigma <- 0.01
mu <- 1*c(1,0.9,0.8,0.4,0.2,0.1)
ndose <- length(mu)-1
dat <- as.data.frame( t(plyr::ldply(mu,function(x){rlnorm(N,x,sigma)})))
names(dat) <- c("C",paste0("T",1:ndose))

dat <- pivot_longer(dat,cols=everything()) %>% mutate(name=factor(name))
plot(dat$value~dat$name)
ggplot(dat,aes(x=name,y=value))+geom_boxplot()+geom_point()
res1 <- williamsTest(dat$value~dat$name,alternative = "less")
summary(res1)
getwilliamRes(res1)
jonckheereTest(dat$value~dat$name,alternative = "less")
kruskalTest(dat$value~dat$name,alternative = "less")
cor.test(dat$value, as.numeric(dat$name), method = "kendall", alternative = "two.sided", continuity = TRUE)


res2 <- stepDownTrendTest(value~name,data=dat,test="jonckheereTest",alternative = "less")
# res2 <- stepDownTrendTest(value~name,data=dat,test="jonckheereTest")
summary(res2)

res3 <- shirleyWilliamsTest(value~name,data=dat,alternative = "less")
summary(res3)
```



```{r}
res <- stepDownTrendTest(Y ~ DOSE, data = trout,
                         test = "jonckheereTest",
                         alternative = "less")
## print method
res
## summary method
summary(res)

williamsTest(Y ~ DOSE, data = trout,alternative = "less")

ggplot(trout,aes(x=DOSE,y=Y))+geom_boxplot()+geom_point()
```



## Control FWER 

| Observed p-values | Bonferroni | Holm Step-down   | Hochberg Step-up |
|-------------------|------------|------------------|------------------|
| $p_{r_1}$         | $\alpha/G$ | $\alpha/G$       | $\alpha/G$       |
| $p_{r_2}$         |            | $\alpha/(G-1)$   | $\alpha/G$       |
| . .               | . .        | . .              | . .              |
| $p_{r_g}$         | $\alpha/G$ | $\alpha/(G-g+1)$ | $\alpha/(G-g+1)$ |
| . .               | . .        | . .              | . .              |
| $p_{r_{G-1}}$     | $\alpha/G$ | $\alpha/2$       | $\alpha/2$       |
| $p_{r_G}$         | $\alpha/G$ | $\alpha$         | $\alpha$         |

## Jonckheere Trend Test

A step‐down procedure based on trend works as
follows: First, the hypothesis that there is no trend in
response with increasing dose is tested when the control
and all dose groups are included in the test. Then, if the test
for trend is significant, the high‐dose group is dropped
from the dataset, and the hypothesis that there is no trend in the reduced dataset is tested. This process of dropping the
highest treatment group and retesting the reduced dataset
is continued until the first time the trend test is not significant.
The highest dose in the reduced dataset at that stage
is then declared to be the NOEC. Distinguishing features of
step‐down procedures are that the tests of hypothesis must
be performed in a given order, and that the outcome of each
hypothesis test is evaluated before deciding whether to test
the next hypothesis in the ordered sequence of hypotheses.
It is these two aspects of these procedures that account for
controlling the FWE rate.(???) Indeed, the special class of “fixed‐sequence” tests that
we will describe fixes the critical level at 0.05 for each
comparison but still bound the FWE rate at 0.05. Thus,
step‐down methods are generally preferable to the singlestep
methods as long as the population response means
are monotonic. (???)

Tests based on trend are logically consistent with the
anticipated monotone pattern of responses in toxicity tests.
Step‐down procedures make use of this ordered alternative
by ordering the tests of hypotheses. This minimizes the
number of comparisons that need to be made.


```{r}

```



```{r}
## Example from Sachs (1997, p. 402)
x <- c(106, 114, 116, 127, 145,
       110, 125, 143, 148, 151,
       136, 139, 149, 160, 174)
g <- gl(3,5)
levels(g) <- c("A", "B", "C")

## Chacko's test
chackoTest(x, g)

## Cuzick's test
cuzickTest(x, g)

## Johnson-Mehrotra test
johnsonTest(x, g)

## Jonckheere-Terpstra test
jonckheereTest(x, g)

## Le's test
leTest(x, g)

## Spearman type test
spearmanTest(x, g)

## Murakami's BWS trend test
bwsTrendTest(x, g)

## Fligner-Wolfe test
flignerWolfeTest(x, g)
```


**References**
- Jonckheere, A. R. (1954) A distribution-free k-sample test against ordered alternatives. Biometrica 41, 133–145.

- Kloke, J., McKean, J. W. (2015) Nonparametric statistical methods using R. Boca Raton, FL: Chapman & Hall/CRC.

## Shirley-Williams Test

```{r}


## Example from Sachs (1997, p. 402)
x <- c(106, 114, 116, 127, 145,
110, 125, 143, 148, 151,
136, 139, 149, 160, 174)
g <- gl(3,5)
levels(g) <- c("0", "I", "II")

## Williams Test
williamsTest(x ~ g)
plot(x~g)

## Example from Shirley (1977)
## Reaction times of mice to stimuli to their tails.
x <- c(2.4, 3, 3, 2.2, 2.2, 2.2, 2.2, 2.8, 2, 3,
 2.8, 2.2, 3.8, 9.4, 8.4, 3, 3.2, 4.4, 3.2, 7.4, 9.8, 3.2, 5.8,
 7.8, 2.6, 2.2, 6.2, 9.4, 7.8, 3.4, 7, 9.8, 9.4, 8.8, 8.8, 3.4,
 9, 8.4, 2.4, 7.8)
g <- gl(4, 10)

plot(x~g)
## Shirley's test
## one-sided test using look-up table
shirleyWilliamsTest(x ~ g, alternative = "greater")

## Chacko's global hypothesis test for 'greater'
chackoTest(x , g)

## post-hoc test, default is standard normal distribution (NPT'-test)
summary(chaAllPairsNashimotoTest(x, g, p.adjust.method = "none"))

## same but h-distribution (NPY'-test)
chaAllPairsNashimotoTest(x, g, dist = "h")

## NPM-test
NPMTest(x, g)

## Hayter-Stone test
hayterStoneTest(x, g)

## all-pairs comparisons
hsAllPairsTest(x, g)
```



## Nonparametric MCP and trend test

- Shirley-type (non-parametric version of Williams Test)
- Other provided by ***nparcomp*


The function **nparcomp** computes the estimator of nonparametric relative contrast effects, simultaneous confidence intervals for the effects and simultaneous p-values based on special contrasts like *"Tukey", "Dunnett", "Sequen", "Williams", "Changepoint", "AVE", "McDermott", "Marcus", "UmbrellaWilliams", "UserDefined"*. The statistics are computed using multivariate normal distribution, multivariate Satterthwaite t-Approximation and multivariate transformations (Probit and Logit transformation function). The function 'nparcomp' also computes one-sided and two-sided confidence intervals and p-values. The confidence intervals can be plotted.

```{r}
library(nparcomp)
data(liver)

  # Williams Contrast

a<-nparcomp(weight ~dosage, data=liver, asy.method = "probit",
            type = "Williams", alternative = "two.sided", 
            plot.simci = TRUE, info = FALSE,correlation=TRUE)
summary(a)
```


```{r}
 # Dunnett dose 3 is baseline

c<-nparcomp(weight ~dosage, data=liver, asy.method = "probit",
            type = "Dunnett", control = "3",
            alternative = "two.sided", info = FALSE)
summary(c)
plot(c)

```


```{r}
data(colu)

  # Tukey comparison- one sided(lower)

a<-nparcomp(corpora~ dose, data=colu, asy.method = "mult.t",
            type = "Tukey",alternative = "less", 
            plot.simci = TRUE, info = FALSE)
summary(a)


 # Tukey comparison- one sided(greater)

b<-nparcomp(corpora~ dose, data=colu, asy.method = "mult.t",
            type = "Tukey",alternative = "greater", 
            plot.simci = TRUE, info = FALSE)
summary(b)
```



## Nonparametric equivalent for two sample t-test

The function npar.t.test performs two sample tests for the nonparametric Behrens-Fisher problem, that is testing the hypothesis

$$H_0: p=1/2,$$

where p denotes the relative effect of 2 independent samples and computes confidence intervals for the relative effect p. The statistics are computed using standard normal distribution, Satterthwaite t-Approximation and variance stabilising transformations (Probit and Logit transformation function). For small samples there is also a studentized permutation test implemented. npar.t.test also computes one-sided and two-sided confidence intervals and p-values. The confidence interval can be plotted.


```{r}
data(impla)
a<-npar.t.test(impla~group, data = impla, method = "t.app",
               alternative = "two.sided", info=FALSE)
summary(a)
plot(a)               
b<-npar.t.test(impla~group, data = impla, method= "permu",
               alternative = "two.sided", info=FALSE)
summary(b)
plot(b)
```


## 

```{r}
b<-mctp(weight ~dosage, data=liver, asy.method = "fisher",
        type = "Dunnett", alternative = "two.sided", 
        plot.simci = TRUE, info = FALSE)
summary(b)
```


## References

- Frank Konietschke \& Ludwig A. Hothorn (2012) Evaluation of Toxicological Studies Using a Nonparametric Shirley-Type Trend Test for Comparing Several Dose Levels with a Control Group, Statistics in Biopharmaceutical Research, 4:1, 14-27, DOI: 10.1080/19466315.2011.633861